Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-03-20 paper.tex
Update paper.tex
Updated section 1 with a description of the chapter

The traditional, reductionist view of biology has considered the
behavior and interactions of individual molecular entities as a way to
understand and explain biological phenomena. This approach has become
embodied in the ``one-target, one-drug'' approach to drug
development. Leaving aside the subtleties of how targets are defined
\cite{Imming:2006lt}, it has become increasingly clear that many drugs
(and more generally, small molecules) exhibit polypharmacological
behavior - interacting with multiple targets, including their putative
primary target. In many cases, the activity against the primary target
is significantly higher than off-target activities, minimizing the
effects of the latter. However, this is not always the case as in the
example of terfenadine (a histamine H1 antagonist), where off target
activity against hERG \cite{Crumb:1995it} led to an increased risk of
cardiac death. Concurrently, it is becoming increasingly clear that
many diseases are not driven by dysfunction of a single
target. Examples include diabetes and cancer. Even when a single
target is the main driver for a given disease state, compensatory
pathways can ``route around'' inhibition of the primary target, thus
restoring the functional behavior that the primary target is
responsible (i.e., the system is now resistant to the original drug)
\cite{Sierra:2010mz,Dienstmann:2012qr}. While the preceding discussion
presents polypharmacological behavior as a negative characteristic,
this is not always the case. Indeed, modern approaches to drug
development now consider such behavior attractive, leading to efforts
to rationally design or control polypharmacology
\cite{Gujral:2014dp,Ciceri:2014fv,Metz:2010ys}.

Given these observations it is clear that pathways play fundamental
roles in biological systems. Naturally, approaches that analyse these
networked systems can be effective in understanding their behavior and
then identifying ways to modulate them. Importantly, network
approaches provide the opportunity to integrate multiple entity types
(small molecules, protein targets, genes) in a systematic fashion.The
remainder of this chapter looks at how network methods can be applied
to various topics in drug discovery ranging from drug-target
interactions to analyzing structure-activity relationships.

@@

Using these databases, network methods can be applied to various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness (as measured by essentiality and centrality) may lead to
significant side effects in other parts of the system and can be the
source of observed side effects \cite{Wang:2013gn}. On these lines,
approaches to characterize toxicity have employed network analyses
\cite{Kim:2014xs,Zhang:2014hj}. While some approaches have considered
side effects [REF] others have considered the problem at the genomic
level by characterizing gene networks that are perturbed by small
molecules. Campaigns focused on combination therapies can also benefit
from a network representation and network inference approaches
\cite{Huang:2014by,Tang:2013lr} to identify target combinations whose
simultaneous inhibition leads to synergistic effects. Mechanism of
action studies can benefit from network analyses by integrating
multiple data types. An example of this was described by Young et al
\cite{Young:2008aa} that integrated high content screening and
ligand-target prediction models to construct network representations
of MoA hypotheses. Network approaches have also been proposed for SAR
analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at the lead
optimization stage of the discovery pipeline. Drug repositioning
efforts \cite{Dudley:2011dn,Gottlieb:2011rx} benefit from network
analyses such as community detection algorithms
\cite{girvan2002community}.

@@

Since small molecules are fundamentally weighted networks (of atoms
and bonds), a variety of graph algorithms can be applied to them to
derive invariants (also known as topological descriptors
\cite{Guha:2012vn}). However, small molecules interact with protein
targets and the targets themselves interact with each other. Such
explicit interactions lend themselves naturally to network
representations. More generally, observed or computed relationships
between small molecules, protein or gene targets and diseases allow
one to develop network representations of these entities which can
then be visualized and quantified to support integrative analyses of
multiple data types. In the following subsections we highlight
applications of this approach and where possible provide examples of R
code to generate and analyze such networks.

@@

Many workers have described analytic approaches based on drug-target
networks. For example, the MetaCore and MetaDrug suite of tools have
been used to perform a variety of functional analyses on small
molecules, particularly metabolites \cite{Brennan:2009kx}.  Examples
include pathway enrichment analyses and pathway reconstruction. The
latter is useful in placing new or predicted metabolites in the
context of known protein-protein and protein-compound
interactions. Iwata et al \cite{Iwata:2013lk} described the use of a
drug-target network to identify protein domains that are associated
with drug side effects. Wang et al \cite{Wang:2013gn} characterized
the topological properties of drug-target networks and noted that
target essentiality and centrality were strongly linked to drug side
effects and that a drugs' side effects were primarily influenced by
the number of \emph{essential} targets rather than the number of drug
targets. Drug-target networks have been used as input to a variety of
predictive models. For example, a model to predict ATC codes from a
drug-target network was described by Wang et al
\cite{Wang:2013ye}. Tabei et al \cite{Tabei:2012tn} used a drug-target
network a input to a classifier that was used to extract chemogenomic
features. These features represent associations between chemical
substructures and protein domains. The set of features that provides a
compressed representation of the drug-target network and the authors
proposed that they could be used to suggest new substructures for a
given protein domain.

@@

@@

We would like to thank Dac-Trung Nguyen and Tyler Peryea for fruitful
discussions related scaffold and scaffold-document networks.

>>>>

The traditional, reductionist view of biology has considered the
behavior and interactions of individual molecular entities as a way to
understand and explain biological phenomena. This approach has become
embodied in the ``one-target, one-drug'' approach to drug
development. Leaving aside the subtleties of how targets are defined
\cite{Imming:2006lt}, it has become increasingly clear that many small
molecules exhibit polypharmacological behavior - interacting with
multiple targets, including their putative primary target. In many
cases, the activity against the primary target is significantly higher
than off-target activities, minimizing the effects of the
latter. However, this is not always the case as in the example of
terfenadine (a histamine H1 antagonist), where off target activity
against hERG \cite{Crumb:1995it} led to an increased risk of cardiac
death. Concurrently, it is becoming increasingly clear that many
diseases are not driven by dysfunction of a single target. Examples
include diabetes and cancer. Even when a single target is the main
driver for a given disease state, compensatory pathways can ``route
around'' inhibition of the primary target, thus restoring the
functional behavior that the primary target is responsible (i.e., the
system is now resistant to the original drug)
\cite{Sierra:2010mz,Dienstmann:2012qr}. While this discussion presents
polypharmacological behavior as a negative characteristic, this is not
always the case. Indeed, modern approaches to drug development now
consider such behavior attractive, leading to efforts to rationally
design or control polypharmacology
\cite{Gujral:2014dp,Ciceri:2014fv,Metz:2010ys}.

Given these observations it is clear that pathways play fundamental
roles in biological systems. Naturally, approaches that analyse these
networked systems can be effective in understanding their behavior and
then identifying ways to modulate them. Importantly, network
approaches provide the opportunity to integrate multiple entity types
(small molecules, protein targets, genes) in a systematic fashion.

The remainder of this chapter looks at how network methods can be
applied to various topics in drug discovery ranging from drug-target
interactions to analyzing structure-activity relationships. In
addition to reviewing methodologies, we also highlight implementations
of these methods and related tools that are available for network
analyses in the R environment. We conclude the chapter with a
discussion of working with chemical structure data in the R
environment and how such structures can be employed in a network
analysis.

@@

Using these databases, network methods can be applied at various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness (as measured by essentiality and centrality) may lead to
significant side effects in other parts of the system and can be the
source of observed side effects \cite{Wang:2013gn}. On these lines,
approaches to characterize toxicity have employed network analyses
\cite{Kim:2014xs,Zhang:2014hj}. While some approaches have considered
side effects [REF] others have considered the problem at the genomic
level by characterizing gene networks that are perturbed by small
molecules. Campaigns focused on combination therapies can also benefit
from a network representation and network inference approaches
\cite{Huang:2014by,Tang:2013lr} to identify target combinations whose
simultaneous inhibition leads to synergistic effects. Mechanism of
action studies can benefit from network analyses by integrating
multiple data types. An example of this was described by Young et al
\cite{Young:2008aa} that integrated high content screening and
ligand-target prediction models to construct network representations
of MoA hypotheses. Network approaches have also been proposed for SAR
analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at the lead
optimization stage. Drug repositioning efforts
\cite{Dudley:2011dn,Gottlieb:2011rx} benefit from network analyses
such as community detection algorithms \cite{girvan2002community}.

@@

Since small molecules are fundamentally weighted networks (of atoms
and bonds), a variety of graph algorithms can be applied to them to
derive invariants (also known as topological descriptors
\cite{Guha:2012vn}). In this work we focus on small molecule
interactions with protein targets and the targets themselves interact
with each other. Such explicit interactions lend themselves naturally
to network representations. More generally, observed or computed
relationships between small molecules, protein or gene targets and
diseases allow one to develop network representations of these
entities which can then be visualized and quantified to support
integrative analyses of multiple data types. In the following
subsections we highlight applications of this approach and where
possible provide examples of R code to generate and analyze such
networks.

@@

Many workers have described analytic approaches based on drug-target
networks. For example, the MetaCore and MetaDrug suite of tools have
been used to perform a variety of functional analyses on small
molecules, particularly metabolites \cite{Brennan:2009kx}.  Examples
include pathway enrichment analyses and pathway reconstruction. The
latter is useful in placing new or predicted metabolites in the
context of known protein-protein and protein-compound
interactions. Iwata et al \cite{Iwata:2013lk} described the use of a
drug-target network to identify protein domains that are associated
with drug side effects. Wang et al \cite{Wang:2013gn} characterized
the topological properties of drug-target networks and noted that
target essentiality and centrality were strongly linked to drug side
effects and that a drugs' side effects were primarily influenced by
the number of \emph{essential} targets rather than the number of drug
targets. Drug-target networks have been used as input to a variety of
predictive models. For example, a model to predict ATC codes from a
drug-target network was described by Wang et al
\cite{Wang:2013ye}. Tabei et al \cite{Tabei:2012tn} used a drug-target
network as input to a classifier that was used to extract chemogenomic
features. These features represent associations between chemical
substructures and protein domains. The set of features that provides a
compressed representation of the drug-target network and the authors
proposed that they could be used to suggest new substructures for a
given protein domain.

@@

A variety of algorithms related to repositioning and activity
predictions, based on drug-target networks have been implemented in
R. Ma and Zhao \cite{Ma:2012qy} implemented the FacPad method that
utilizes a Bayesian sparse factor model to identify pathways that are
perturbed upon drug treatments. The DrugComboRanker package
\cite{Huang:2014by} employed a community based partitioning to create
a drug meta-community network and a disease signaling meta-community
network. Drug targets were then mapped onto disease networks to allow
identification of synergistic pairs of drugs and complementary
mechanisms of action. In the tool called MANTRA (Mode of Action by
NeTwoRk Analysis) \cite{iorio2010discovery}, authors used network
communities to deduce mechanism of action of new molecules and suggest
repositioning of a well known molecule to a new therapeutic
arean. Finally the TIMMA-R package \cite{He:2015ly} combines binary
drug-target data and drug-sensitivity profiles to predict synergistic
pairs of anti-cancer drugs.

@@

We would like to thank Dac-Trung Nguyen and Tyler Peryea for fruitful
discussions related to scaffold and scaffold-document networks.
